FinalManuscript_DISCS_UG

### 2019-04-29 parts/08_appendices_b_numericalapproximations_updated.tex
Fixed graphs in appendix
Final revisions and cleanup

\chapter{LOGARITHM APPROXIMATION}
\label{chap:logarithm_approximation}
Many methods to compute transcendental functions implemented in computer hardware and software libraries rely on direct knowledge of the function arguments. Several such approaches include CORDIC (coordinate rotation digital computer), pseudo-division and pseudo-multiplication algorithms~\cite{walther_cordic_2000}, as well as minimax polynomials, which are determined numerically using the Remez algorithm~\cite{harrison_computation_1999}. These algorithms, while efficient, rely on table lookups, which cannot be done in a privacy-preserving system where the function arguments cannot be directly determined.

We must therefore rely on other methods, such as using the Taylor series expansions of these functions. While also in common use, not all Taylor series expansions have a convenient radius of convergence. To circumvent this, we can rely on an approach presented by Khattri in~\cite{khattri_new_2009} to obtain closed-form approximations to functions by approximating related integrals.

@@

While the closed-form approximation presented by Khattri (Equation~\ref{eq:standard_logarithm_quadrature}) is accurate for values of $x$ near zero, it diverges from $\log{\left(1+x\right)}$ significantly for large values of $x$, as shown in Figure~\ref{fig:standard_logarithm_quadrature}.
\begin{figure}[!ht]
	\centering
	% \begin{tikzpicture}
	% 	\begin{axis}[
	% 		% height=6cm, width=9cm,
	% 		xmin=0, xmax=300,
	% 		ymin=0, ymax=5.5, ytick={0,...,6},
	% 		xlabel={$x$}, ylabel={$y$},
	% 		grid=major,
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 	]
	% 	\addplot[very thick, blue, domain=0:300, samples=100]{1/30*(137*x^5 + 2310*x^4 + 9870*x^3 + 15120*x^2 + 7560*x)/(x^5 + 30*x^4 + 210*x^3 + 560*x^2 + 630*x + 252)};
	% 	\addplot[densely dashed, domain=0:300, samples=100]{ln(x)};
	% 	\legend{5-point quadrature approximation, Actual value of $\log(1+x)$}
	% 	\end{axis}
	% \end{tikzpicture}%
	\caption{Graph of $\log{\left(1+x\right)}$ and the approximation in Equation~\ref{eq:standard_logarithm_quadrature}}
	\label{fig:standard_logarithm_quadrature}
\end{figure}

@@

\section{Scaled Approximation}
\label{sec:logarithm_approximation}
This appendix a general method to improve the general accuracy of the logarithm approximation given by Khattri in~\cite{khattri_new_2009} for use in privacy-preserving calculation, provided that accuracy is only required for a specific range.

@@

\begin{figure}[ht]
	\centering
  %   \begin{tikzpicture}
	% 	\begin{semilogyaxis}[
	% 		% height=6cm, width=9cm,
	% 		xmin=0, xmax=300,
	% 		ymin=1e-10, ymax=1.4,
	% 		xlabel={$x$}, ylabel={Absolute error},
	% 		grid=major,
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 		tick scale binop=\times,
	% 	]
	% 	% \addplot[very thick, domain=0:300, samples=500]{abs(-1/30*(x^5*(30*ln(1/16) - 137) + 45*x^4*(270*ln(1/16) - 593) + 30*x^3*(27210*ln(1/16) - 20579) + 30*x^2*(488810*ln(1/16) + 216621) + 45*x*(1690670*ln(1/16) + 2694207) + 93677310*ln(1/16) + 257804775)/(x^5 + 405*x^4 + 27210*x^3 + 488810*x^2 + 2536005*x + 3122577) - ln(x + 1))};
	% 	% \addplot[very thick, domain=0:300, samples=300]{abs(1/30*(137*x^5 + 26685*x^4 + 617370*x^3 - 6498630*x^2 - 121239315*x - 257804775)/(x^5 + 405*x^4 + 27210*x^3 + 488810*x^2 + 2536005*x + 3122577) - ln(1/16) - ln(x + 1))};
	% 	\addplot[very thick, blue] table[x=x, y=error] {plots/abs_err_scaledlog.txt};
	% 	\addplot[densely dashed, domain=0:300, samples=100]{abs(1/30*(137*x^5 + 2310*x^4 + 9870*x^3 + 15120*x^2 + 7560*x)/(x^5 + 30*x^4 + 210*x^3 + 560*x^2 + 630*x + 252) - ln(x + 1))};
	% 	\legend{Scaled quadrature ($\alpha = 1/16$), Original quadrature ($\alpha = 1$)}
	% \end{semilogyaxis}
	% \end{tikzpicture}%
	\caption{Comparison of the absolute error between the original approximation in Equation~\ref{eq:standard_logarithm_quadrature} and the scaled approximation in Equation~\ref{eq:optimal_log_approximation}}
	\label{fig:log_error_comparison}
\end{figure}

@@

Let $I\left(x,\alpha\right)$ be the function obtained by approximating $\int_{n}^{\alpha n + \alpha}{\frac{1}{x}\diff x}$ with five-point Gauss--Legendre quadrature. The absolute error of the approximation compared to $\log{\left(1+x\right)}$ at a given $x$ and $\alpha$ is given by the following function.
\begin{align*}
	A\left(x,\alpha\right) = \left| I\left(x,\alpha\right) - \log\alpha - \log{\left(1+x\right)}\right|.
\end{align*}
We want to show that the function
\begin{align*}
	f\left(\alpha\right) = \max_{0\leq x \leq 255}{A\left(x,\alpha\right)}
\end{align*}
is minimized at $\alpha = 1/16$.
We consider the functions $A\left(0,\alpha\right)$ and $A\left(255,\alpha\right)$. These are plotted in Figure~\ref{fig:endpoint_plot}.
\begin{figure}[!ht]
	\centering
  %   \begin{tikzpicture}
	% 	\begin{semilogyaxis}[
	% 		% height=6cm, width=9cm,
	% 		xmin=0, xmax=1,
	% 		ymin=5e-7, ymax=1.3,
	% 		xlabel={$\alpha$}, ylabel={Absolute error},
	% 		grid=major,
	% 		extra x ticks={1/16},
	% 		extra x tick labels={$\frac{1}{16}$},
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 		tick scale binop=\times,
	% 	]
	% 	\addplot[thick, ACMDarkBlue, domain=0:1] table[x=alpha, y=x0] {plots/abs_err_sl_0_255.txt};
	% 	\addplot[thick, ACMOrange, domain=0:1] table[x=alpha, y=x255] {plots/abs_err_sl_0_255.txt};
	% 	\legend{$x=0$, $x=255$}
	% 	% \legend{$A\left(0,\alpha \right)$, $A\left(255,\alpha \right)$}
	% \end{semilogyaxis}
	% \end{tikzpicture}%
	\caption{Graphs of $A\left(x,\alpha\right)$ at $x=0$ and $x=255$}
	\label{fig:endpoint_plot}
\end{figure}

@@

We then consider the function $A\left(x,1/16\right)$, shown in Figure~\ref{fig:single_alpha_plot}.
It was determined numerically that the local maxima of $A\left(x,1/16\right)$ on the interval occur at $x=0$ and $x=255$. Since a scaling factor other than $\alpha=1/16$ yields a higher absolute error at either $x=0$ or $x=255$, as seen in Figure~\ref{fig:endpoint_plot}, it follows that a scaling factor of $\alpha = 1/16$ minimizes the maximum absolute error of the scaled approximation for $\log{\left(1+x\right)}$, within the range $x \in [0, 255]$.
\begin{figure}[ht]
	\centering
	% \begin{tikzpicture}
	% 	\begin{semilogyaxis}[
	% 		xmin=0, xmax=300,
	% 		ymin=1e-15, ymax=0.035,
	% 		xlabel={$x$}, ylabel={Absolute error},
	% 		grid=major,
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 		tick scale binop=\times,
	% 	]
	% 	% \addplot[very thick, domain=0:300, samples=300]{abs(1/30*(137*x^5 + 26685*x^4 + 617370*x^3 - 6498630*x^2 - 121239315*x - 257804775)/(x^5 + 405*x^4 + 27210*x^3 + 488810*x^2 + 2536005*x + 3122577) - ln(1/16) - ln(x + 1))};
	% 	\addplot[thick, ACMDarkBlue] table[x=x, y=error] {plots/abs_err_scaledlog.txt};
	% \end{semilogyaxis}
	% \end{tikzpicture}%
	\caption{Graph of $A\left(x,1/16\right)$, the absolute error of the approximation in Equation~\ref{eq:optimal_log_approximation} given a scaling factor of $\alpha=1/16$}
	\label{fig:single_alpha_plot}
\end{figure}

@@

\section{Approximation for $x^\gamma$}
To approximate $x^\gamma$ for any $\gamma \in \mathbb{R}$, we rewrite $x^\gamma$ as follows:
\begin{align*}
  x^\gamma = e^{\log{x^\gamma}} = e^{\gamma\log{x}}.
\end{align*}
This expression can then be approximated using the Maclaurin series expansion for $e^x$, which converges for all $x$.
\begin{align*}
  e^x &= \sum_{n=0}^{\infty}{\frac{x^n}{n!}}\\
  \Rightarrow e^{\gamma\log{x}} &= \sum_{n=0}^{\infty}{\frac{(\gamma\log{x})^n}{n!}}
\end{align*}
As we already have an approximation for the natural logarithm, we can evaluate partial sums of the above infinite series to arrive at approximations for $x^\gamma$.

>>>>

\chapter{PRIVACY-PRESERVING LOGARITHM APPROXIMATION}
\label{chap:logarithm_approximation}
Many methods to compute transcendental functions implemented in computer hardware and software libraries rely on direct knowledge of the function arguments. Several such approaches include CORDIC (coordinate rotation digital computer), pseudo-division and pseudo-multiplication algorithms~\cite{walther_cordic_2000}, as well as minimax polynomials, which are determined numerically using the Remez algorithm~\cite{harrison_computation_1999}. These algorithms, while efficient, rely on table lookups, which cannot be done in a privacy-preserving system where the function arguments cannot be directly determined. We must rely on other methods such as series expansions.

While in common use, not all series expansions have a convenient radius of convergence. To circumvent this, we can rely on an approach presented by Khattri in~\cite{khattri_new_2009} to obtain closed-form approximations to functions by approximating related integrals.

@@

While the closed-form approximation presented by Khattri (Equation~\ref{eq:standard_logarithm_quadrature}) is accurate for values of $x$ near zero, it diverges from $\log{\left(1+x\right)}$ significantly for large values of $x$, as shown in Figure~\ref{fig:standard_logarithm_quadrature}.
\begin{figure}[!ht]
	\centering
	% \begin{tikzpicture}
	% 	\begin{axis}[
	% 		% height=6cm, width=9cm,
	% 		xmin=0, xmax=300,
	% 		ymin=0, ymax=5.5, ytick={0,...,6},
	% 		xlabel={$x$}, ylabel={$y$},
	% 		grid=major,
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 	]
	% 	\addplot[very thick, blue, domain=0:300, samples=100]{1/30*(137*x^5 + 2310*x^4 + 9870*x^3 + 15120*x^2 + 7560*x)/(x^5 + 30*x^4 + 210*x^3 + 560*x^2 + 630*x + 252)};
	% 	\addplot[densely dashed, domain=0:300, samples=100]{ln(x)};
	% 	\legend{5-point quadrature approximation, Actual value of $\log(1+x)$}
	% 	\end{axis}
	% \end{tikzpicture}%
    \includegraphics[width=12cm]{figures/graph_log_a2.png}
	\caption{Graph of $\log{\left(1+x\right)}$ and the approximation in Equation~\ref{eq:standard_logarithm_quadrature}}
	\label{fig:standard_logarithm_quadrature}
\end{figure}

@@

\section{Scaled Approximation}
\label{sec:logarithm_approximation}
We now present a method to improve the general accuracy of the logarithm approximation given by Khattri in~\cite{khattri_new_2009} for use in privacy-preserving calculation, provided that accuracy is only required for a specific range.

@@

\begin{figure}[ht]
	\centering
  %   \begin{tikzpicture}
	% 	\begin{semilogyaxis}[
	% 		% height=6cm, width=9cm,
	% 		xmin=0, xmax=300,
	% 		ymin=1e-10, ymax=1.4,
	% 		xlabel={$x$}, ylabel={Absolute error},
	% 		grid=major,
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 		tick scale binop=\times,
	% 	]
	% 	% \addplot[very thick, domain=0:300, samples=500]{abs(-1/30*(x^5*(30*ln(1/16) - 137) + 45*x^4*(270*ln(1/16) - 593) + 30*x^3*(27210*ln(1/16) - 20579) + 30*x^2*(488810*ln(1/16) + 216621) + 45*x*(1690670*ln(1/16) + 2694207) + 93677310*ln(1/16) + 257804775)/(x^5 + 405*x^4 + 27210*x^3 + 488810*x^2 + 2536005*x + 3122577) - ln(x + 1))};
	% 	% \addplot[very thick, domain=0:300, samples=300]{abs(1/30*(137*x^5 + 26685*x^4 + 617370*x^3 - 6498630*x^2 - 121239315*x - 257804775)/(x^5 + 405*x^4 + 27210*x^3 + 488810*x^2 + 2536005*x + 3122577) - ln(1/16) - ln(x + 1))};
	% 	\addplot[very thick, blue] table[x=x, y=error] {plots/abs_err_scaledlog.txt};
	% 	\addplot[densely dashed, domain=0:300, samples=100]{abs(1/30*(137*x^5 + 2310*x^4 + 9870*x^3 + 15120*x^2 + 7560*x)/(x^5 + 30*x^4 + 210*x^3 + 560*x^2 + 630*x + 252) - ln(x + 1))};
	% 	\legend{Scaled quadrature ($\alpha = 1/16$), Original quadrature ($\alpha = 1$)}
	% \end{semilogyaxis}
	% \end{tikzpicture}%
    \includegraphics[width=12cm]{figures/comp_errs.png}
	\caption{Comparison of the absolute error between the original approximation in Equation~\ref{eq:standard_logarithm_quadrature} and the scaled approximation in Equation~\ref{eq:optimal_log_approximation}}
	\label{fig:log_error_comparison}
\end{figure}

@@

Let $I\left(x,\alpha\right)$ be the function obtained by approximating $\int_{n}^{\alpha n + \alpha}{\frac{1}{x}\diff x}$ with five-point Gauss--Legendre quadrature. The absolute error of the approximation compared to $\log{\left(1+x\right)}$ at a given $x$ and $\alpha$ is given by the following function.
\begin{align*}
	A\left(x,\alpha\right) = \left| I\left(x,\alpha\right) - \log\alpha - \log{\left(1+x\right)}\right|.
\end{align*}
We want to show that the function
\begin{align*}
	f\left(\alpha\right) = \max_{0\leq x \leq 255}{A\left(x,\alpha\right)}
\end{align*}
is minimized at $\alpha = 1/16$.
We consider the functions $A\left(0,\alpha\right)$ and $A\left(255,\alpha\right)$. These are plotted in Figure~\ref{fig:endpoint_plot}.
\begin{figure}[!ht]
	\centering
  %   \begin{tikzpicture}
	% 	\begin{semilogyaxis}[
	% 		% height=6cm, width=9cm,
	% 		xmin=0, xmax=1,
	% 		ymin=5e-7, ymax=1.3,
	% 		xlabel={$\alpha$}, ylabel={Absolute error},
	% 		grid=major,
	% 		extra x ticks={1/16},
	% 		extra x tick labels={$\frac{1}{16}$},
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 		tick scale binop=\times,
	% 	]
	% 	\addplot[thick, ACMDarkBlue, domain=0:1] table[x=alpha, y=x0] {plots/abs_err_sl_0_255.txt};
	% 	\addplot[thick, ACMOrange, domain=0:1] table[x=alpha, y=x255] {plots/abs_err_sl_0_255.txt};
	% 	\legend{$x=0$, $x=255$}
	% 	% \legend{$A\left(0,\alpha \right)$, $A\left(255,\alpha \right)$}
	% \end{semilogyaxis}
	% \end{tikzpicture}%
    \includegraphics[width=12cm]{figures/graph_0_255.png}
	\caption{Graphs of $A\left(x,\alpha\right)$ at $x=0$ and $x=255$}
	\label{fig:endpoint_plot}
\end{figure}

@@

We then consider the function $A\left(x,1/16\right)$, shown in Figure~\ref{fig:single_alpha_plot}.
It was determined numerically that the local maxima of $A\left(x,1/16\right)$ on the interval occur at $x=0$ and $x=255$. Since a scaling factor other than $\alpha=1/16$ yields a higher absolute error at either $x=0$ or $x=255$, as seen in Figure~\ref{fig:endpoint_plot}, it follows that a scaling factor of $\alpha = 1/16$ minimizes the maximum absolute error of the scaled approximation for $\log{\left(1+x\right)}$, within the range $x \in [0, 255]$.
\begin{figure}[ht]
	\centering
	% \begin{tikzpicture}
	% 	\begin{semilogyaxis}[
	% 		xmin=0, xmax=300,
	% 		ymin=1e-15, ymax=0.035,
	% 		xlabel={$x$}, ylabel={Absolute error},
	% 		grid=major,
	% 		legend pos=south east,
	% 		legend cell align=left,
	% 		tick scale binop=\times,
	% 	]
	% 	% \addplot[very thick, domain=0:300, samples=300]{abs(1/30*(137*x^5 + 26685*x^4 + 617370*x^3 - 6498630*x^2 - 121239315*x - 257804775)/(x^5 + 405*x^4 + 27210*x^3 + 488810*x^2 + 2536005*x + 3122577) - ln(1/16) - ln(x + 1))};
	% 	\addplot[thick, ACMDarkBlue] table[x=x, y=error] {plots/abs_err_scaledlog.txt};
	% \end{semilogyaxis}
	% \end{tikzpicture}%
    \includegraphics[width=12cm]{figures/graph_approx_16.png}
	\caption{Graph of $A\left(x,1/16\right)$, the absolute error of the approximation in Equation~\ref{eq:optimal_log_approximation} given a scaling factor of $\alpha=1/16$}
	\label{fig:single_alpha_plot}
\end{figure}

@@

\section{Approximation for $x^\gamma$}
\label{sec:power_approximation}
To approximate $x^\gamma$ for any $\gamma \in \mathbb{R}$, we rewrite $x^\gamma$ as follows:
\begin{align*}
  x^\gamma = e^{\log{x^\gamma}} = e^{\gamma\log{x}}.
\end{align*}
This expression can then be approximated using the Maclaurin series expansion for $e^x$, which converges for all $x$.
\begin{align*}
  e^x &= \sum_{n=0}^{\infty}{\frac{x^n}{n!}}
  &e^{\gamma\log{x}} = \sum_{n=0}^{\infty}{\frac{(\gamma\log{x})^n}{n!}}
\end{align*}
As we already have an approximation for the natural logarithm, we can evaluate partial sums of the above infinite series to arrive at approximations for $x^\gamma$.
